---
title: "Reproducible Research: Peer Assessment 1"
author: "Hanna Zhao"
date: "August 15, 2016"
output: html_document;md_document

---

## Loading and preprocessing the data

First step: load "activity.csv" file from R working directory and assign it to a data frame named "activity". 
It has three columns:  
*- steps: Number of steps taking in a 5-minute interval  
- date: The date on which the measurement was taken in YYYY-MM-DD format  
- interval: Identifier for the 5-minute interval in which measurement was taken*

```{r loaddata, echo=TRUE}
activity<-read.csv("D:/Data Scientist/R_working_directory/activity.csv",na.strings="NA")
head(activity)
```

Convert "date" column to Date class by "as.Date" function.

```{r convert, echo=TRUE}
activity$date<-as.Date(activity$date,"%Y-%m-%d")
```


## What is mean total number of steps taken per day?

Calculate total number of steps per day.

```{r simulation, echo=TRUE}
AC_by_day<-aggregate(activity$steps,by=list(activity$date),FUN="sum",na.rm=TRUE)
names(AC_by_day)<-c("date","total.steps")
```

Here is the histogram of the total number of steps per day.
```{r, echo=TRUE}
library(ggplot2)
g<-qplot(total.steps,data=AC_by_day,binwidth=500)
g+labs(title="Total Steps per Day")
#dev.copy(png,"plot1.png")
#dev.off()
```

Mean of total daily steps is:
```{r,echo=TRUE}
StpMean<-mean(AC_by_day$total.steps,na.rm=TRUE)
StpMean
```

Median of total daily steps is:
```{r,echo=TRUE}
StpMedian<-median(AC_by_day$total.steps,na.rm=TRUE)
StpMedian
```

## What is the average daily activity pattern?
Calculate average steps at each 5 mins interval order across all days. Plot the average steps vs interval order.
```{r,echo=TRUE}
AC_by_Int<-aggregate(activity$steps,by=list(activity$interval),FUN="mean",na.rm=TRUE)
names(AC_by_Int)<-c("Interval","steps")

qplot(AC_by_Int$Interval,AC_by_Int$steps,
         geom=c("line"),
         xlab="Interval Order",ylab="Average steps across all days"
      )

#dev.copy(png,"plot2.png")
#dev.off()
```

The maximum steps in 5-min interval is:
```{r,echo=TRUE}
IntStepMax<-AC_by_Int[which.max(AC_by_Int$steps),1]
IntStepMax
```


## Imputing missing values
Original dataset contains many missing data points. Total number of NA rows is:
```{r,echo=TRUE}
sum(is.na(activity$steps))
```

Now let's add missing values by using the mean for that 5-min interval and save it to a new data frame named "acivity2".
```{r,echo=TRUE}
ind_NA<-which(is.na(activity$steps)==TRUE)
activity2<-activity

for (i in ind_NA) {
      activity2$steps[i]<-AC_by_Int$steps[AC_by_Int$Interval==activity2$interval[i]]
}
```

Check the missing values in "activity2". 
```{r,echo=TRUE}
sum(is.na(activity2))
```

Plot the histogram of the total number of steps per day. the new data frame "activity2" after adding missing values. 
```{r.echo=TRUE}
AC_by_day2<-aggregate(activity2$steps,by=list(activity2$date),FUN="sum",na.rm=TRUE)
names(AC_by_day2)<-c("date","total.steps")

g<-qplot(total.steps,data=AC_by_day2,binwidth=500)
g+labs(title="Total Steps per Day after Adding Missing Values")

#dev.copy(png,"plot3.png")
#dev.off()
```

Compare it with the distogram from original data. Before adding missing values, the distogram of original data has a lot of actificial zero points due to missing values. The mean and median value of total daily steps calculated from original dataset are also misleading due to this issue. 

Mean of total daily steps after adding missing values.
```{r,echo=TRUE}
StpMean2<-mean(AC_by_day2$total.steps,na.rm=TRUE)
StpMean2
```

Median of total daily steps after adding missing values.
```{r,echo=TRUE}
StpMedian2<-median(AC_by_day2$total.steps,na.rm=TRUE)
StpMedian2
```

## Are there differences in activity patterns between weekdays and weekends?
First add weekday and weekend information to "activity2" data frame. Then calculate the average steps at each 5-min interval for each category.
```{r,echo=TRUE}
tmp<-weekdays(activity2$date)
tmp2<-sapply(tmp,function(x) {
              if (x=="Saturday"|x=="Sunday") "Weekend"
              else   "Weekday"
})

activity2<-cbind(activity,tmp,tmp2)
names(activity2)[4:5]<-c("Weekday.name","Weekday")

AC_by_Int2<-aggregate(activity2$steps,by=list(activity2$Weekday,activity$interval),FUN="mean",na.rm=TRUE)
names(AC_by_Int2)<-c("Weekday","Interval","steps")
AC_by_Int2$Weekday<-as.factor(AC_by_Int2$Weekday)
```

Plot the activity pattern for both weekdays and weekends.
```{r,echo=TRUE}
g<-ggplot(AC_by_Int2,aes(Interval,steps))
g+geom_line()+facet_grid(Weekday~.)

#dev.copy(png,"plot4.png")
#dev.off()
```


